### 2023-10-02 appendices/paper_brief.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses

\paragraph{Moving forward} 
Most of the seminal work has focused on improving speech recognition \parencite{schneider_wav2vec_2019, baevski_wav2vec_2020}. This focus has gained traction over the last couple of years, as computational resources have become more accessible and end-to-end models \parencite{graves_connectionist_2006, chan_listen_2015} have been established as the dominant approach to speech recognition \parencite{gulati_conformer_2020}. It is important to stress that self-supervised models, such as wav2vec 2.0 \parencite{baevski_wav2vec_2020}, represent a breakthrough, and recent successful approaches build upon this method. That is, deep self-attention models combined with masking \parencite{hsu_hubert_2021, wang_unispeech_2021, chen_wavlm_2021}. This development mirrors years of rapid progress in masked language modeling within natural language processing \parencite{devlin_bert_2018,clark_electra_2020} and we expect this to continue for unsupervised neural speech representation learning.

>>>>

\paragraph{Moving forward} 
Most of the seminal work has focused on improving speech recognition \parencite{schneider_wav2vec_2019, baevski_wav2vec_2020}. This focus has gained traction over the last couple of years, as computational resources have become more accessible and end-to-end models \parencite{graves_connectionist_2006, chan_listen_2016} have been established as the dominant approach to speech recognition \parencite{gulati_conformer_2020}. It is important to stress that self-supervised models, such as wav2vec 2.0 \parencite{baevski_wav2vec_2020}, represent a breakthrough, and recent successful approaches build upon this method. That is, deep self-attention models combined with masking \parencite{hsu_hubert_2021, wang_unispeech_2021, chen_wavlm_2021}. This development mirrors years of rapid progress in masked language modeling within natural language processing \parencite{devlin_bert_2018,clark_electra_2020} and we expect this to continue for unsupervised neural speech representation learning.

### 2023-10-02 chapters/background.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses

Other methods note that the difficulty of detecting OOD data might be attributed to the curse of dimensionality in the learned feature spaces and propose to use dimensionality reduction techniques. In \textcite{ndiour_outofdistribution_2020}, apply dimensionality reduction on learned, high-dimensional features to capture the true feature subspace and compute the norm of the difference between the original feature and the pre-image of its low-dimensional manifold embedding. 
\textcite{zaeemzadeh_outofdistribution_2021} forces the ID samples to embed into a union of 1-dimensional subspaces during training and computes the minimum angular distance from the feature to the class-wise subspaces. 
NuSA \parencite{cook_outlier_2020} uses projects features onto the column space of the classification weight matrix and computes the ratio of the norm the projected and original features.
\textcite{lee_simple_2018} fit a multivariate Gaussian distribution to the activations of the penultimate layer of a pre-trained classifier and use the Mahalanobis distance to this distribution to evaluate whether inputs are OOD. This method can also be seen as ameliorating the curse of dimensionality by clustering the high-dimensional feature space. 
Finally, Bayesian neural networks have also been proposed for OOD detection although their performance is not yet competitive with other methods \parencite{henning_are_2021,dangelo_outofdistribution_2022,nguyen_out_2022}.

@@

\paragraph{Reparameterization} 
Via the ELBO, we want to use maximum likelihood estimation to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$. 
Gradient-based optimization works well for neural networks but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
Computing the gradient of the ELBO with respect to $\thetab$ is straightforward,
%
\begin{equation}
    % \nabla_\thetab \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] 
    \nabla_\thetab \mathcal{L}_{\thetab,\phib}(\xb)
    = \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \nabla_\thetab \log p_\thetab(\xb|\zb) \right] \enspace .
\end{equation}
%
The gradient with respect to the $\phib$ is more challenging since $\phib$ occurs in the expectation itself making it impossible to move the gradient inside the expectation. 
However, by reparameterizing the latent variable $\zb$ as a deterministic function of the input $\xb$ and a random variable $\epsilonb\sim p(\epsilonb)$, $\zb = g_\phib(\xb,\epsilonb)$, we can write the gradient as a path-wise derivative,
%
\begin{equation}
    % \nabla_\phib \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] 
    \nabla_\phib \mathcal{L}_{\thetab,\phib}(\xb)
    = \mathbb{E}_{p(\epsilonb)} \left[ \nabla_\phib \log \frac{p_\thetab\left(\xb|g_\phib(\xb,\epsilonb)\right)p\left(g_\phib(\xb,\epsilonb)\right)}{q_\phib\left(g_\phib(\xb,\epsilonb)|\xb\right)} \right] \enspace . \\
\end{equation}
%
The most important instantiation of this reparameterization is the diagonal Gaussian case, where $\zb = \mu_\phib(\xb) + \sigma_\phib(\xb) \odot \epsilonb$ with $\epsilonb\sim\mathcal{N}(0,\mathbf{I})$. It is usual to make a mean-field approximation by assuming the covariance matrix is diagonal imposing independence between the latent variables in $\zb$.

Although lower variance than alternatives, such as score function estimators, a number of works have attempted to reduce the variance. 
\textcite{roeder_sticking_2017} note that the path-wise derivative can be factored into a path-wise estimator and a score function estimator and propose to ignore the score function factor to reduce gradient variance. 
\textcite{rainforth_tighter_2019} that this problem is even larger for the importance weighted ELBO \parencite{burda_importance_2016} leading to diminishing returns for increasing numbers of samples. Instead of removing the score function factor in the gradient, \textcite{tucker_doubly_2019} propose to reparameterize the score function and derive low variance gradient estimators also for importance weighted ELBO.

@@

However, this alternative form reveals that the ELBO is maximized by minizing the mutual information between $\xb$ and $\zb$, $I_{q_{}(\xb,\zb)}\left[\xb ; \zb \right]$. 

@@

% \paragraph{Hierarchical models}
% VAEs can be extended to hierarchical models by introducing additional latent variables $\zb_1,\dots,\zb_L$ that are conditionally dependent on each other \parencite{ranganath_hierarchical_2016,sonderby_ladder_2016,vahdat_nvae_2020,child_very_2021,maaloe_biva_2019}. The generative model can then be written as,
% %
% \begin{equation}
%     p_\thetab(\xb,\zb_1,\dots,\zb_L) = p_\thetab(\xb|\zb_1)\prod_{l=2}^L p_\thetab(\zb_{l-1}|\zb_{l}) \enspace .
% \end{equation}
% %

>>>>

Other methods note that the difficulty of detecting OOD data might be attributed to the curse of dimensionality in the learned feature spaces and propose to use dimensionality reduction techniques. In \textcite{ndiour_outofdistribution_2020}, apply dimensionality reduction on learned, high-dimensional features to capture the true feature subspace and compute the norm of the difference between the original feature and the pre-image of its low-dimensional manifold embedding. 
\textcite{zaeemzadeh_outofdistribution_2021} forces the ID samples to embed into a union of 1-dimensional subspaces during training and computes the minimum angular distance from the feature to the class-wise subspaces. 
NuSA \parencite{cook_outlier_2020} uses projects features onto the column space of the classification weight matrix and computes the ratio of the norm the projected and original features.
\textcite{lee_simple_2018} fit a multivariate Gaussian distribution to the activations of the penultimate layer of a pre-trained classifier and use the Mahalanobis distance to this distribution to evaluate whether inputs are OOD. This method can also be seen as ameliorating the curse of dimensionality by clustering the high-dimensional feature space. 
Finally, Bayesian neural networks have also been proposed for OOD detection, but their performance is not yet competitive with other methods \parencite{henning_are_2021,dangelo_outofdistribution_2022,nguyen_out_2022}.

@@

\paragraph{Reparameterization} 
Via the ELBO, we want to use maximum likelihood estimation to jointly optimize the generative model $p_\thetab(\xb|\zb)$ and the variational approximation $q_\phib(\zb|\xb)$ with respect to $\thetab$ and $\phib$. 
Gradient-based optimization works well for neural networks but the large amounts of data force the use of mini-batch, or stochastic, gradient descent. 
Computing the gradient of the ELBO with respect to $\thetab$ is straightforward,
%
\begin{equation}
    % \nabla_\thetab \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] 
    \nabla_\thetab \mathcal{L}_{\thetab,\phib}(\xb)
    = \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \nabla_\thetab \log p_\thetab(\xb|\zb) \right] \enspace .
\end{equation}
%
The gradient with respect to the $\phib$ is more challenging since $\phib$ occurs in the expectation itself making it impossible to move the gradient inside the expectation. 
However, by reparameterizing the latent variable $\zb$ as a deterministic function of the input $\xb$ and a random variable $\epsilonb\sim p(\epsilonb)$, $\zb = g_\phib(\xb,\epsilonb)$, we can write the gradient as a path-wise derivative,
%
\begin{equation}
    % \nabla_\phib \mathbb{E}_{q_\phib(\zb|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p(\zb)}{q_\phib(\zb|\xb)} \right] 
    \nabla_\phib \mathcal{L}_{\thetab,\phib}(\xb)
    = \mathbb{E}_{p(\epsilonb)} \left[ \nabla_\phib \log \frac{p_\thetab\left(\xb|g_\phib(\xb,\epsilonb)\right)p\left(g_\phib(\xb,\epsilonb)\right)}{q_\phib\left(g_\phib(\xb,\epsilonb)|\xb\right)} \right] \enspace . \\
\end{equation}
%
The most important instantiation of this reparameterization is the diagonal Gaussian case, where $\zb = \mu_\phib(\xb) + \sigma_\phib(\xb) \odot \epsilonb$ with $\epsilonb\sim\mathcal{N}(0,\mathbf{I})$. 
By assuming the covariance matrix is diagonal, this is a mean-field approximation which imposes independence between the latent variables in $\zb$.

Although the path-wise derivative has lower variance than most alternatives, such as score function estimators, a number of works have attempted to reduce its variance.
\textcite{roeder_sticking_2017} note that the path-wise derivative can be factored into a path-wise estimator and a score function estimator and propose to ignore the score function factor to reduce gradient variance. 
\textcite{rainforth_tighter_2019} show that this problem is even larger for the importance weighted ELBO \parencite{burda_importance_2016} and leads to diminishing returns when increasing the number of samples. 
Rather than removing the score function factor in the estimate, \textcite{tucker_doubly_2019} propose to reparameterize it and derive low variance gradient estimators, also for importance weighted ELBO.

\paragraph{Hierarchical models}\label{sec:hierarchical-vae-background}
The usual mean-field assumption imposed by letting $\zb \sim\mathcal{N}(\mub,\mathbf{I}\sigmab^2)$ can often be more restrictive for model expressiveness than we would like. 
While learning a full covariance is a simple solution, it is not always sufficiently computationally efficient and, in any case, enables learning only linear covariance between elements of $\zb$. 
Along with the usual motivations of hierarchical models such as efficient representation, feature disentanglement, and data efficiency, this has led to research into using a hierarchy of several conditionally dependent latent variables. 
This idea is commonly formalized by introducing a set of $L$ latent variables  $\zb=\zb_1, \dots, \zb_L$ and letting the generative model be defined as,
%
\begin{equation}
    p_\thetab(\xb|\zb)=p(\xb|\zb_1) p_\thetab(\zb_1|\zb_2) \cdots p_\thetab(\zb_{L-1}|\zb_L) p(\zb_L) \enspace .
\end{equation}
%
This top-down generative model can be efficiently sampled via ancestral sampling; first $\zb_L$ is drawn from the prior $p(\zb_L)$, and then each $\zb_l$ is drawn from the corresponding $p(\zb_l|\zb_{l+1})$ until we can draw $p(\xb|\zb_1)$.

The inference model can then be defined in two ways respectively referred to as \emph{bottom-up} \parencite{burda_importance_2016}
\begin{equation}
    %q_\phib(\zb|\xb)=q_\phib(\zb_1|\xb)q_\phib(\zb_2|\zb_1)\cdots q_\phib(\zb_L|\zb_{L-1})
    q_\phib(\zb|\xb) = q_\phib(\zb_1|\xb)\textstyle\prod_{i=2}^{L} q_\phib(\zb_i|\zb_{i-1})
\end{equation}
and \emph{top-down} \parencite{sonderby_ladder_2016}
\begin{equation}
    % q_\phib(\zb|\xb)=q_\phib(\zb_{L}|\xb)q_\phib(\zb_{L-1}|\zb_L)\cdots q_\phib(\zb_2|\zb_1),
    q_\phib(\zb|\xb) = q_\phib(\zb_L|\xb)\textstyle\prod_{i=L-1}^{1} q_\phib(\zb_{i}|\zb_{i+1}) \ .
\end{equation}
Regardless of the choice of inference model, the resulting hierarchical VAE is trained using the ELBO \cref{eq:variational-autoencoder-elbo} and the reparameterization trick.

% Until recently, hierarchical VAEs gave inferior likelihoods compared to state-of-the-art autoregressive \parencite{ho_flow_2019} and flow-based models \parencite{salimans_pixelcnn_2017}.
% This was changed by \textcite{maaloe_biva_2019}, \textcite{vahdat_nvae_2020}, and \textcite{child_very_2021}, which introduced complementary methods to extend the number of latent variables to a very deep hierarchy resulting in state-of-the-art likelihood performance.

@@

However, this alternative form reveals that the ELBO is maximized by minimizing the mutual information between $\xb$ and $\zb$, $I_{q_{}(\xb,\zb)}\left[\xb ; \zb \right]$. 

@@

% %

### 2023-10-02 chapters/discussion.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses

\chapter[discussion]{Discussion}\label{chp:discussion}
% ~5 pages

@@

@@

\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation? \parencite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.}
\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation?}
% Since the main focus within self-supervised learning has been on improving downstream task performance, very limited work, if any, has investigated self-supervised representations in terms of uncertainty estimation. 
% However, in the context of medical applications where data can be abundant but labels are sparse, unsupervised uncertainty estimation is a very interesting direction for future work.
\lesstodo[inline]{OOD data: Generalization or detection (https://arxiv.org/pdf/2110.11334.pdf)?}

\section{\Cref{chp:paper-retrospective} revisited: Uncertainty}
The work in \textcite{wenstrup_retrospective_2023} focuses on predictive performance and an analysis of feature importance but does not explicitly consider uncertainty estimation. To

@@

\lesstodo[inline]{Discuss the calibration of the stroke model and report calibration curve.}
\lesstodo[inline]{Report the results of calibrating the stroke model using e.g. Platt scaling or Isotonic scaling.} % https://scikit-learn.org/stable/modules/calibration.html
\lesstodo[inline]{Maybe mention MultiQT paper \parencite{havtorn_multiqt_2020}.}

>>>>

\chapter[discussion]{Discussion}\label{chp:discussion}
% ~5 pages
%
% OUTLINE:
% - paper-hierarchical:
%   - Suitability of VAEs for representation learning (minimization of mutual information and sensitivity to implicit prior such as architecture)
% - paper-benchmarking
%   - Inferiority of probabilistic methods compared to self-supervised learning. 
% -

@@

\section{\Cref{chp:paper-benchmarking} revisited: Bested?}

\lesstodo[inline]{Discuss whether variational autoencoders are viable for learning good representations for speech for downstream tasks when alternatives such as SSL method exist.}
\lesstodo[inline]{Out-of-distribution detection on speech?}

@@

\lesstodo[inline]{Are self-supervised speech representations useful for unsupervised uncertainty estimation? \parencite{nava_stateconsistency_2021, nava_uncertaintyaware_2021} Within robotics but not really related.}
% Since the main focus within self-supervised learning has been on improving downstream task performance, very limited work, if any, has investigated self-supervised representations in terms of uncertainty estimation. 
% However, in the context of medical applications where data can be abundant but labels are sparse, unsupervised uncertainty estimation is a very interesting direction for future work.
\lesstodo[inline]{OOD data: Generalization or detection (https://arxiv.org/pdf/2110.11334.pdf)?}

\section{\Cref{chp:paper-retrospective} revisited: Uncertainty}
The work in \textcite{wenstrup_retrospective_2023} focuses on predictive performance and an analysis of feature importance but does not explicitly consider uncertainty estimation.

@@

\lesstodo[inline]{Discuss the calibration of the stroke model and report calibration curve.}
\lesstodo[inline]{Report the results of calibrating the stroke model using e.g. Platt scaling or Isotonic scaling.} % https://scikit-learn.org/stable/modules/calibration.html

% \lesstodo[inline]{Maybe mention MultiQT paper \parencite{havtorn_multiqt_2020}.}

### 2023-10-02 chapters/introduction.tex
Add hierarchical VAE section and update main contributions with research questions/hypotheses

\section{Thesis outline and research questions}

@@

This introduction provided a high-level overview of the motivation behind the research project by focusing on the use-cases of stroke recognition and automated medical coding as well as speech processing and the importance of uncertainty quantification. 
The remainder of \cref{part:background} consists of two chapters. 
\Cref{chp:technical-background} provides relevant technical background to the papers of the thesis, including uncertainty, out-of-distribution detection, and variational autoencoders while \cref{chp:main-contributions} outlines the main contributions of the thesis via the published papers in relation to the research hypotheses.

\Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. 
\Cref{part:unsupervised-uncertainty-estimation} contains papers on variational autoencoders and their applications to out-of-distribution detection and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}. Here we examine whether it is possible to devise a method that utilizes a hierarhy of latent variables in

\Cref{part:self-supervised-speech-representation-learning} contains a review paper on self-supervised speech representation learning \parencite{mohamed_selfsupervised_2022}. \Cref{part:medical-applications} contains papers on applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}.

@@

\section{Thesis scope and outline}
%
% OUTLINE: 
% - Tie together the different topics of the thesis and point briefly to the relevant papers.
%
% The complete set of research published as part of this project covers a number of topics related to decision-support in medical conversations. 
% One group of the primary papers is concerned with generative latent variable models and their applications to estimation of epistemic uncertainty and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}. 
% Another group of studies examines self-supervised speech representation learning as an alternative to generative latent variable models \parencite{mohamed_selfsupervised_2022,borgholt_brief_2022} and explores them for speech recognition \parencite{borgholt_scaling_2021} and spoken language understanding \parencite{borgholt_we_2021}. 
% Finally, a group deals with applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}. 
%

>>>>

\section{Thesis scope and outline}

@@

This introduction provided a high-level overview of the motivation behind the research project by focusing on the use-cases of stroke recognition and automated medical coding as well as speech processing and the importance of uncertainty quantification. 
The remainder of \cref{part:background} consists of two chapters. 
\Cref{chp:main-contributions} outlines the research questions of the thesis via the published papers and presents their main contributions. 
\Cref{chp:technical-background} provides relevant technical background to the papers of the thesis, including uncertainty, out-of-distribution detection, and variational autoencoders.

\Cref{part:unsupervised-uncertainty-estimation,part:self-supervised-speech-representation-learning,part:medical-applications} contain a number of chapters that each correspond to a primary paper. 
\Cref{part:unsupervised-uncertainty-estimation} contains papers on variational autoencoders and their applications to out-of-distribution detection and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}. 
\Cref{part:self-supervised-speech-representation-learning} contains a review paper on self-supervised speech representation learning \parencite{mohamed_selfsupervised_2022}. \Cref{part:medical-applications} contains papers on applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}.

@@

% \section{Thesis scope and outline}
%
% OUTLINE: 
% - Tie together the different topics of the thesis and point briefly to the relevant papers.
%
% The complete set of research published as part of this project covers a number of topics related to decision-support in medical conversations. 
% One group of the primary papers is concerned with generative latent variable models and their applications to estimation of epistemic uncertainty and speech modelling \parencite{havtorn_hierarchical_2021,havtorn_benchmarking_2022,bergamin_modelagnostic_2022}. 
% Another group of studies examines self-supervised speech representation learning as an alternative to generative latent variable models \parencite{mohamed_selfsupervised_2022,borgholt_brief_2022} and explores them for speech recognition \parencite{borgholt_scaling_2021} and spoken language understanding \parencite{borgholt_we_2021}. 
% Finally, a group deals with applications of machine learning within the medical domain, including recognition of stroke cases in calls to medical helplines \parencite{wenstrup_retrospective_2023} and medical coding of clinical notes \parencite{edin_automated_2023}. 
%
